---
title: "SAS Codes"
author: "Harry Lu 呂達"
output:
  html_document:
    toc: TRUE
    toc_float: 
      collapsed: FALSE
    toc_depth: 2
    number_sections: TRUE
---
<style>
h1 {font-size: 28px; color: black; font-weight: bold; text-align: left; background-color: Orange;}
h1.title {color: black; text-align: center; background-color: White;}
h2 {font-size: 20px; color: black; font-weight: bold; font-style: italic; background-color: Yellow;}
h4.author {color: black; text-align: center;}
p {font-size: 16px; color: black;}
ol {font-size: 16px; color: black;}
ul {font-size: 16px; color: black;}
pre {margin-left: 40px;}
table {margin-left: 40px; font-size: 16px;}
</style>

---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
saspath = "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts = "-nosplash -ls 75"
```

```{r, engine='sas', engine.path=saspath, engine.opts=sasopts, comment=NA, eval=FALSE}
/*==============================Problem 1==============================*/
/*
- Create a library called HW1 with the dataset Admit
*/
libname HW1 "C:/Users/Harry Lu's X1/Desktop/Cornell/STSCI 5010/HW 1";
title1 "Summary of Admit Dataset";
footnote1 "Produced by Da (Harry) Lu";
OPTIONS NONUMBER NODATE PAGESIZE = 20;
PROC CONTENTS data = HW1.Admit;
RUN;
/*
- There are 21 observations.
- There are 9 Variables: ID, Name, Sex, Age, Date, Height, Weight, ActLevel, Fee.
- The dataset was created on 09/19/2013 at 18:22:16.
- There are 4 Char variables: ActLevel, ID, Name, Sex.
- There are 5 Num variables: Age, Date, Fee, Height, Weight.
*/

/*==============================Problem 2==============================*/
/*
- Create al_high
*/
title1 "Dataset al_high";
DATA al_high;
	Set HW1.Admit;
	Where ActLevel = "HIGH";
RUN;
PROC PRINT data = al_high;
RUN;
/*
- Create al_low
*/
title1 "Dataset al_low";
DATA al_low;
	Set HW1.Admit;
	Where ActLevel = "LOW";
RUN;
PROC PRINT data = al_low;
RUN;
/*
- Create al_mod
*/
title1 "Dataset al_mod";
DATA al_mod;
	Set HW1.Admit;
	Where ActLevel = "MOD";
RUN;
PROC PRINT data = al_mod;
RUN;
/*
- Without adding exact lib directory to the front of the newly created dataset
	the dataset will be automatically saved to Work Library. However, after closing
	the current session, the three new dataset will not be permanantly saved at
	Work Library so we cannot find it, unless we re-run the program to re-create it.
*/

/*==============================Problem 3==============================*/
/*----------------------------Problem 3 (a)----------------------------*/
title1 "Patients with HIGH or MOD ActLevel and  pay a Fee within
the range of [100, 130]";
PROC PRINT data = HW1.Admit;
	WHERE (ActLevel = "HIGH" OR ActLevel = "MOD") AND (Fee >= 100 AND Fee <= 130);
RUN;

/*----------------------------Problem 3 (b)----------------------------*/
title1 "Patients whose names contains 'J' and 'R'";
PROC PRINT data = HW1.Admit;
	VAR ID Name Sex Age;
	WHERE Name CONTAINS "J" and Name CONTAINS "R";
RUN;

/*----------------------------Problem 3 (c)----------------------------*/
title1 "Female patients who pay a fee that is greater than $100";
Data c;
	Set HW1.Admit;
	Label ActLevel = "Activity Level";
	Format Fee DOLLAR8.2;
RUN;
PROC PRINT data = c label;
	ID ID;
	WHERE Sex = "F" AND Fee > 100;
RUN;

/*==============================Problem 4==============================*/
title1 "Male patients who are at least 65 years old";
Data HW1.OldMale;
	Set HW1.Admit;
	Label Fee = "Fee charged at the time of admission ($)";
	WHERE Sex = "M" AND Age >= 65;
	Format Height 6.2;
RUN;
PROC PRINT data=HW1.OldMale Label;
RUN;
```


#OPTIONS
- Syntax

```{}
OPTIONS options;
```

- `DATE`|`NODATE` to display or suppress dates on the output
- `NUMBER`|`NONUMBER` to display or suppress page #
- `PAGENO=` to specify the beginning page number
- `PAGESIZE=` (or `PS=`) to specify the number of line per page of output
- `LINESIZE=` (or `LS=`) to specify the width of the printed line
- `YEARCUTOFF=` to specify which 100-year span is used to interpret two-digit year vales
- `FIRSTOBS=` to start processing at a specific observation
- `OBS=` to stop processing after a specific observation

- Example

```{}
options nonumber nodate;
PROC PRINT data=clinic. payrollmaster;
run;
```

- Example

```{}
options pageno=1 pagesize=15;
PROC PRINT data=clinic. payrollmaster;
run;
```

- Example (Global)

```{}
options firstobs=50 obs=100;
PROC PRINT data=clinic. payrollmaster;
run;
```

- Example (Local)

```{}
PROC PRINT data=clinic. payrollmaster
(firstobs=50 obs=100);
run;
```

- Example

```{}
options yearcutoff=1950;
```

##Titles and Footnotes
- Syntax

```{}
TITLE<n> 'text';
FOOTNOTE<n> 'text';
```

- Example

```{}
options pagesize=20 nodate;
title1 'Heart Rates for Patients with';
title3 'Increased Stress Tolerance Levels';
footnote1 'Data from Treadmill Tests';
footnote3 '1st Quarter Admissions';
proc print data=clinic.stress2;
  var resthr maxhr rechr;
  where tolerance='I';
run;
```





#Comments in SAS
- Syntax

```{}
* Prints the data;
```

- Example

```{}
proc print data= student;
run;
```

- Syntax

```{}
/* Prints the data*/
```

- Example

```{}
proc print data= students;
run;
```



#PROC
##PRINT
```{}
PROC PRINT data = xxxxxxxx;
RUN;
```

**SUM**

- Example

```{}
proc print data=clinic.insure;
  var name policy balancedue;
  where pctinsured < 100;
  sum balancedue;
run; 
```

**SUM + BY**

- Example

```{}
* The data set must be sorted first;
proc sort data=clinic.admit out=work.activity;
  by actlevel;
run;
proc print data=work.activity;
  var age height weight fee;
  where age>30;
  sum fee;
  by actlevel;
run; 
```

**SUM + BY + ID**

- Example

```{}
* The data set must be sorted first;
proc sort data=clinic.admit out=work.activity;
  by actlevel;
run;
proc print data=work.activity;
  var age height weight fee;
  where age>30;
  sum fee;
  by actlevel;
  id actlevel;
run;
```

**SUM + BY + ID + PAGEBY**

- Example

```{}
* The data set must be sorted first;
proc sort data=clinic.admit out=work.activity;
  by actlevel;
run;
proc print data=work.activity;
  var age height weight fee;
  where age>30;
  sum fee;
  by actlevel;
  id actlevel;
  pageby actlevel;
run;
```

**DOUBLE**

- Example

```{}
proc print data=clinic.stress2 double;
  var resthr maxhr rechr;
  where tolerance='I';
run;
```

**LABEL**

- Example

```{}
proc print data=clinic.admit label;
  var age height;
  label age='Age of Patient';
  label height='Height in Inches';
run;
```

- Example

```{}
proc print data=clinic.admit label;
  var age height;
  label age='Age of Patient'
    height='Height in Inches';
run;
```

**FORMAT**

Format|Specifies values ...|Example
:----------------|:--------------------------------|:----------------
COMMAw.d|that contain commas and decimal places|comma8.2
DOLLAR w.d|that contain dollar signs, commas, and decimal places|dollar6.2
MMDDYYw.|as date values of the form 09/12/97 (MMDDYY8.) or 09/12/1997 (MMDDYY10.)|mmddyy10.
w.|rounded to the nearest integer in w spaces|7.
w.d|rounded to d decimal places in w spaces|8.2
\$w.|as character values in w spaces|\$12.
DATEw.|as date values of the form 16OCT99 (DATE7.) or 16OCT1999 (DATE9.)|date9.

- Example

```{}
proc print data=clinic.admit;
  var actlevel fee;
  where actlevel='HIGH';
  format fee dollar4.;
run;
```

Examples of SAS Formats, Stored Values and Displayed Values

Stored Value|Format|Displayed Value
:-|:-|:-
38245.3975|COMMA12.2|38,245.40
38245.3975|12.2|38245.40
38245.3975|DOLLAR12.2|\$38,245.40
38245.3975|DOLLAR9.2|\$38245.40
38245.3975|DOLLAR8.2|38245.40
0|MMDDYY8.|01/01/60
0|MMDDYY10.|01/01/1960
0|DATE7.|01JAN60
0|DATE9.|01JAN1960

**Permanently Assigned Labels and Formats**

- [click](SAS_Codes.html#41_permanently_assigned_labels_and_formats)

##CONTENTS
**Summary of dataset**

- Syntax

```{}
PROC CONTENTS<options>;
RUN;
```

- Example

```{}
proc contents data=clinic._all_nods;
run;
```

- Syntax

```{}
PROC DATASETS<options>;
CONTENTS<options>;
QUIT;
```

- Example

```{}
proc datasets;
contents data=clinic._all_nods;
quit;
```

##VARNUM
- Example

```{}
proc contents data=clinic.admit varnum;
run;
```

```{}
proc datasets;
contents data=clinic.admit varnum;
quit;
```

##SORT
- Syntax

```{}
PROC SORT DATA=SAS-data-set<OUT=SAS-data-set>;
  BY <DESCENDING> BY-variable(s);
RUN;
```

- Example 1

```{}
proc sort data=clinic.admit out=work.wgtadmit;
  by weight age;
run;
proc print data=work.wgtadmit;
  var weight age height fee;
  where age>30;
run;
```

- Example 2 (DESCENDING option)

```{}
* The data set must be sorted first;
proc sort data=clinic.admit out=work.wgtadmit;
  by descending weight age;
run;
proc print data=work.wgtadmit;
  var weight age height fee;
  where age>30;
run; 
```



#DATA
- Example

```{r, engine='sas', engine.path=saspath, engine.opts=sasopts, comment=NA, eval=FALSE}
DATA temp;
  x = 0;
RUN;
PROC PRINT data = temp;
RUN;
```

##Permanently Assigned Labels and Formats
```{}
data sasuser.paris;
  set sasuser.laguardia;
  where dest="PAR" and (boarded=155 or boarded=146);
  label date=’Departure Date’;
  format date date9.;
run;

proc print data=sasuser.paris label;
  var date dest boarded;
run;
```

#Creating List Reports
##Basic
- Example

```{}
options pagesize=500 linesize=200;
PROC PRINT data = sashelp.cars;
run;
```

By default:

- All observations and variables are printed
- They appear in the same order as in the data set
- Observation numbers appear in the far left

##VAR
- Example

```{}
PROC PRINT data = sashelp.cars;
var make model weight mpg_city;
run;
```

##WHERE
- Example

```{}
Options linesize=80;
PROC PRINT data = sashelp.cars;
var make model weight mpg_city;
where weight>5000;
run;
```

- **comparison operators**

Symbol|Meaning|Example
:-|:-|:-
= or eq|equal to|where name='Jones, C.';
^= or ne|not equal to|where temp ne 212;
\> or gt|greater than|where income\>20000;
\< or lt|less than|where partno lt "BG05";
\>= or ge|greater than or equal to|where id\>='1543';
\<= or le|less than or equal to|where pulse le 85;

- **compound expression**

Operator|Meaning
:----------------|:----------------------------------------------------------------
AND &|and, both. If both expressions are true, then the compound expression is true.
OR \||or, either. If either expression is true, then the compound expression is true.

- Example

```{}
PROC PRINT data = sashelp.cars;
  var make model weight mpg_city;
  where weight>3500 AND make ="Audi";
run;
```

- Example

```{}
PROC PRINT data = sashelp.cars;
  var make model weight mpg_city;
  where weight>3500 AND (make ="Audi" OR make="BMW");
run;
```

**IN**

- Example

```{}
PROC PRINT data = sashelp.cars;
  var make model weight mpg_city;
  where weight>3500 AND make IN ("Audi", "BMW");
run;
```

**CONTAINS (?)**

- Example

```{}
PROC PRINT data = sashelp.cars;
  var make model weight mpg_city;
  where weight>3500 AND (make ? "Au" OR make CONTAINS "BM");
run;
```

**NOOBS**

- Example

```{}
Options linesize=80;
  PROC PRINT data = sashelp.cars noobs;
  var make model weight mpg_city;
  where weight>5000;
run;
```

**ID**

- Example

```{}
Options linesize=80;
  PROC PRINT data = sashelp.cars;
  id horsepower;
  var make model weight mpg_city;
  where weight>5000;
run;
```

**ID**, more variables

- Example

```{}
Options linesize=80;
  PROC PRINT data = sasuser.customer;
  id customer_id customer_name;
  where country='CA';
run;
```


