---
title: "SAS Codes"
author: "Harry Lu 呂達"
output:
  html_document:
    toc: TRUE
    toc_float: 
      collapsed: FALSE
    toc_depth: 2
    number_sections: TRUE
---
<style>
h1 {font-size: 28px; color: black; font-weight: bold; text-align: left; background-color: Orange;}
h1.title {color: black; text-align: center; background-color: White;}
h2 {font-size: 20px; color: black; font-weight: bold; font-style: italic; background-color: Yellow;}
h4.author {color: black; text-align: center;}
p {font-size: 16px; color: black;}
ol {font-size: 16px; color: black;}
ul {font-size: 16px; color: black;}
pre {margin-left: 40px;}
table {margin-left: 40px; font-size: 16px;}
</style>

---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
saspath = "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts = "-nosplash -ls 75"
```

#OPTIONS
- Syntax

```{}
OPTIONS options;
```

- `DATE`|`NODATE` to display or suppress dates on the output
- `NUMBER`|`NONUMBER` to display or suppress page #
- `PAGENO=` to specify the beginning page number
- `PAGESIZE=` (or `PS=`) to specify the number of line per page of output
- `LINESIZE=` (or `LS=`) to specify the width of the printed line
- `YEARCUTOFF=` to specify which 100-year span is used to interpret two-digit year vales
- `FIRSTOBS=` to start processing at a specific observation
- `OBS=` to stop processing after a specific observation

- Example

```{}
options nonumber nodate;
PROC PRINT data=clinic. payrollmaster;
run;
```

- Example

```{}
options pageno=1 pagesize=15;
PROC PRINT data=clinic. payrollmaster;
run;
```

- Example (Global)

```{}
options firstobs=50 obs=100;
PROC PRINT data=clinic. payrollmaster;
run;
```

- Example (Local)

```{}
PROC PRINT data=clinic. payrollmaster
(firstobs=50 obs=100);
run;
```

- Example

```{}
options yearcutoff=1950;
```

##Titles and Footnotes
- Syntax

```{}
TITLE<n> 'text';
FOOTNOTE<n> 'text';
```

- Example

```{}
options pagesize=20 nodate;
title1 'Heart Rates for Patients with';
title3 'Increased Stress Tolerance Levels';
footnote1 'Data from Treadmill Tests';
footnote3 '1st Quarter Admissions';
proc print data=clinic.stress2;
  var resthr maxhr rechr;
  where tolerance='I';
run;
```





#Comments in SAS
- Syntax

```{}
* Prints the data;
```

- Example

```{}
proc print data= student;
run;
```

- Syntax

```{}
/* Prints the data*/
```

- Example

```{}
proc print data= students;
run;
```



#PROC
##PRINT
```{}
PROC PRINT data = xxxxxxxx;
RUN;
```

**SUM**

- Example

```{}
proc print data=clinic.insure;
  var name policy balancedue;
  where pctinsured < 100;
  sum balancedue;
run; 
```

**SUM + BY**

- Example

```{}
* The data set must be sorted first;
proc sort data=clinic.admit out=work.activity;
  by actlevel;
run;
proc print data=work.activity;
  var age height weight fee;
  where age>30;
  sum fee;
  by actlevel;
run; 
```

**SUM + BY + ID**

- Example

```{}
* The data set must be sorted first;
proc sort data=clinic.admit out=work.activity;
  by actlevel;
run;
proc print data=work.activity;
  var age height weight fee;
  where age>30;
  sum fee;
  by actlevel;
  id actlevel;
run;
```

**SUM + BY + ID + PAGEBY**

- Example

```{}
* The data set must be sorted first;
proc sort data=clinic.admit out=work.activity;
  by actlevel;
run;
proc print data=work.activity;
  var age height weight fee;
  where age>30;
  sum fee;
  by actlevel;
  id actlevel;
  pageby actlevel;
run;
```

**DOUBLE**

- Example

```{}
proc print data=clinic.stress2 double;
  var resthr maxhr rechr;
  where tolerance='I';
run;
```

**LABEL**

- Example

```{}
proc print data=clinic.admit label;
  var age height;
  label age='Age of Patient';
  label height='Height in Inches';
run;
```

- Example

```{}
proc print data=clinic.admit label;
  var age height;
  label age='Age of Patient'
    height='Height in Inches';
run;
```

**FORMAT**

Format|Specifies values ...|Example
:----------------|:--------------------------------|:----------------
COMMAw.d|that contain commas and decimal places|comma8.2
DOLLAR w.d|that contain dollar signs, commas, and decimal places|dollar6.2
MMDDYYw.|as date values of the form 09/12/97 (MMDDYY8.) or 09/12/1997 (MMDDYY10.)|mmddyy10.
w.|rounded to the nearest integer in w spaces|7.
w.d|rounded to d decimal places in w spaces|8.2
\$w.|as character values in w spaces|\$12.
DATEw.|as date values of the form 16OCT99 (DATE7.) or 16OCT1999 (DATE9.)|date9.

- Example

```{}
proc print data=clinic.admit;
  var actlevel fee;
  where actlevel='HIGH';
  format fee dollar4.;
run;
```

Examples of SAS Formats, Stored Values and Displayed Values

Stored Value|Format|Displayed Value
:-|:-|:-
38245.3975|COMMA12.2|38,245.40
38245.3975|12.2|38245.40
38245.3975|DOLLAR12.2|\$38,245.40
38245.3975|DOLLAR9.2|\$38245.40
38245.3975|DOLLAR8.2|38245.40
0|MMDDYY8.|01/01/60
0|MMDDYY10.|01/01/1960
0|DATE7.|01JAN60
0|DATE9.|01JAN1960

**Permanently Assigned Labels and Formats**

- [click](SAS_Codes.html#41_permanently_assigned_labels_and_formats)

##CONTENTS
**Summary of dataset**

- Syntax

```{}
PROC CONTENTS<options>;
RUN;
```

- Example

```{}
proc contents data=clinic._all_nods;
run;
```

- Syntax

```{}
PROC DATASETS<options>;
CONTENTS<options>;
QUIT;
```

- Example

```{}
proc datasets;
contents data=clinic._all_nods;
quit;
```

##VARNUM
- Example

```{}
proc contents data=clinic.admit varnum;
run;
```

```{}
proc datasets;
contents data=clinic.admit varnum;
quit;
```

##SORT
- Syntax

```{}
PROC SORT DATA=SAS-data-set<OUT=SAS-data-set>;
  BY <DESCENDING> BY-variable(s);
RUN;
```

- Example 1

```{}
proc sort data=clinic.admit out=work.wgtadmit;
  by weight age;
run;
proc print data=work.wgtadmit;
  var weight age height fee;
  where age>30;
run;
```

- Example 2 (DESCENDING option)

```{}
* The data set must be sorted first;
proc sort data=clinic.admit out=work.wgtadmit;
  by descending weight age;
run;
proc print data=work.wgtadmit;
  var weight age height fee;
  where age>30;
run; 
```



#DATA
- Example

```{r, engine='sas', engine.path=saspath, engine.opts=sasopts, comment=NA}
DATA temp;
  x = 0;
RUN;
PROC PRINT data = temp;
RUN;
```

##Permanently Assigned Labels and Formats
```{}
data sasuser.paris;
  set sasuser.laguardia;
  where dest="PAR" and (boarded=155 or boarded=146);
  label date=’Departure Date’;
  format date date9.;
run;

proc print data=sasuser.paris label;
  var date dest boarded;
run;
```

#Creating List Reports
##Basic
- Example

```{}
options pagesize=500 linesize=200;
PROC PRINT data = sashelp.cars;
run;
```

By default:

- All observations and variables are printed
- They appear in the same order as in the data set
- Observation numbers appear in the far left

##VAR
- Example

```{}
PROC PRINT data = sashelp.cars;
var make model weight mpg_city;
run;
```

##WHERE
- Example

```{}
Options linesize=80;
PROC PRINT data = sashelp.cars;
var make model weight mpg_city;
where weight>5000;
run;
```

- **comparison operators**

Symbol|Meaning|Example
:-|:-|:-
= or eq|equal to|where name='Jones, C.';
^= or ne|not equal to|where temp ne 212;
\> or gt|greater than|where income\>20000;
\< or lt|less than|where partno lt "BG05";
\>= or ge|greater than or equal to|where id\>='1543';
\<= or le|less than or equal to|where pulse le 85;

- **compound expression**

Operator|Meaning
:----------------|:----------------------------------------------------------------
AND &|and, both. If both expressions are true, then the compound expression is true.
OR \||or, either. If either expression is true, then the compound expression is true.

- Example

```{}
PROC PRINT data = sashelp.cars;
  var make model weight mpg_city;
  where weight>3500 AND make ="Audi";
run;
```

- Example

```{}
PROC PRINT data = sashelp.cars;
  var make model weight mpg_city;
  where weight>3500 AND (make ="Audi" OR make="BMW");
run;
```

**IN**

- Example

```{}
PROC PRINT data = sashelp.cars;
  var make model weight mpg_city;
  where weight>3500 AND make IN ("Audi", "BMW");
run;
```

**CONTAINS (?)**

- Example

```{}
PROC PRINT data = sashelp.cars;
  var make model weight mpg_city;
  where weight>3500 AND (make ? "Au" OR make CONTAINS "BM");
run;
```

**NOOBS**

- Example

```{}
Options linesize=80;
  PROC PRINT data = sashelp.cars noobs;
  var make model weight mpg_city;
  where weight>5000;
run;
```

**ID**

- Example

```{}
Options linesize=80;
  PROC PRINT data = sashelp.cars;
  id horsepower;
  var make model weight mpg_city;
  where weight>5000;
run;
```

**ID**, more variables

- Example

```{}
Options linesize=80;
  PROC PRINT data = sasuser.customer;
  id customer_id customer_name;
  where country='CA';
run;
```


