---
title: <center><font face="Baskerville Old Face">**Forecasting Stock Value**</font></center>
output: html_document
---
<div id="backhome"><a href="index.html"><i class="fa fa-home" aria-hidden="true"></i></a></div>
<div id="back"><a href="Projects.html"><i class="fa fa-undo" aria-hidden="true"></i></a></div>

#<font face="Baskerville Old Face">Background</font>

<font face="Times New Roman" size=5>

- Instructor: [Hong-Kun Zhang](http://people.math.umass.edu/~hongkun/), Department of Mathematics and Statistics, University of Massachusetts Amherst
- The research started from May 2016 to July 2016. The original data used was [***Straits Times Index (^STI)***](https://finance.yahoo.com/quote/%5ESTI?p=^STI) from [Yahoo Finance](https://finance.yahoo.com/).
- The historical data for Straits Times Index is no longer available, so [***Alphabet Inc. (GOOG)***](https://finance.yahoo.com/quote/GOOG?p=GOOG) is used in this research note.
- The data was downloaded on January 15, 2018. The data is from August 19, 2004 to January 12, 2018.

</font>

#<font face="Baskerville Old Face">Model: AR(1) - GARCH(1,1)</font>

<font face="Times New Roman" size=5>

$$\hat{r}_{t}=\mu_{t}+\sigma_{t}\cdot\epsilon_{t}$$
We have:

- $r_{t}$: Log Return of Adjusted Close Value
- $\mu_{t}$: Rolling Average of Log Returns
- $\sigma_{t}$: Standard Deviation of Log Returns
- $\epsilon_{t}=\cfrac{r_{t}-\mu_{t}}{\sigma_{t}}$

AR(1)
$$\hat\mu_{t}=a_{t-1}+b_{t-1}\cdot\mu_{t-1}$$

GARCH(1,1)
$${\hat\sigma_{t}}^2=\omega_{t-1}+\alpha_{t-1}\cdot{\mu_{t-1}}^2+\beta_{t-1}\cdot{\sigma_{t-1}}^2$$

Therefore,
$$\hat{r}_{t}=\hat\mu_{t}+\hat\sigma_{t}\cdot\epsilon_{t}=(a_{t-1}+b_{t-1}\cdot\mu_{t-1})+\sqrt{\omega_{t-1}+\alpha_{t-1}\cdot{\mu_{t-1}}^2+\beta_{t-1}\cdot{\sigma_{t-1}}^2}\cdot{\epsilon_{t}}$$

```{r Read_Data_Online,echo=FALSE,eval=FALSE}
#read data online
goog=read.csv("https://raw.githubusercontent.com/HarryLu95/dalu0511/master/Project%20Files/FM%20Research/GOOG.csv")
```

```{r Read_Local_Data,echo=TRUE,eval=TRUE}
#read local data
goog=read.csv("C:/Users/thinkpad/Documents/GitHub/dalu0511/Project Files/FM Research/GOOG.csv")
```

The original data has 7 columns: `Date`, `Open`, `High`, `Low`, `Close`, `Adj.Close`, and `Volumn`.

`Day` and `Log.Return` ($r_t$) are added to the data frame.
$$r_t=\ln(\frac{P_t}{P_{t-1}})$$ where $P$ represents adjusted close value.

```{r Load_Packages,echo=TRUE,message=FALSE,warning=FALSE}
#load package: fGarch
library(fGarch)
```

```{r Basic_Visulization,echo=TRUE,fig.align="center"}
#plot the adjusted value
plot.ts(goog$Adj.Close,ylab="Adjusted Value",main="Alphabet Inc. Adjusted Stock Value from Aug/19/2004 to Jan/12/2018")
plot.ts(diff(goog$Adj.Close),main="Differencing Level 1")
```

```{r Compute_Log_Return,echo=TRUE}
#compute log return value
l=dim(goog)[1]
goog$Day=1:l
goog$Log.Return=NA
for (i in 2:l){
  goog$Log.Return[i]=log(goog$Adj.Close[i]/goog$Adj.Close[i-1],base=exp(1))
}
```

```{r Log_Return_Visulization,echo=TRUE,fig.align="center"}
#plot the log return value
plot.ts(goog$Log.Return,ylab="Log Return",main="Alphabet Inc. Log Return")
```

The first $n$ data (`g1`) were used for Model, and the rest (`g2`) were used for back testing.

```{r Split_data,echo=TRUE}
#divide the date into two groups, the second part is used for back testing
n=as.integer(2800)
#n=floor(l/6*5)
goog$g1=NA
goog$g2=NA
goog$g1[2:n]=goog$Log.Return[2:n]
goog$g2[(n+1):l]=goog$Log.Return[(n+1):l]
```

Using linear model on Rolling `Log.Return` to compute $a$ and $b$, and then $\hat\mu$.

```{r Compute_mu_hat,echo=TRUE,fig.align="center"}
#compute the value of mu hat
goog$mu.hat=NA
for (i in n:(l-1)){
  lm=lm(goog$Log.Return[2:i]~goog$Day[2:i])
  a=lm$coefficient[[1]]
  b=lm$coefficient[[2]]
  mu=mean(goog$Log.Return[2:i])
  temp.mu.hat=a+b*mu
  goog$mu.hat[i+1]=temp.mu.hat
}
#plot.ts(goog$mu.hat[(n+1):l],main="mu hat",ylab="")
```

Using GARCH model on Rolling `Log.Return` to compute $\omega$, $\alpha$ and $\beta$, and then $\hat\sigma^2$.

```{r Compute_sigma_hat_square,echo=TRUE,eval=TRUE}
#compute the value of sigma hat square
goog$sigma.hat.square=NA
for (i in n:(l-1)){
  g=garchFit(~garch(1,1),goog$Log.Return[2:i],trace=FALSE)
  omega=g@fit$coef[[2]]
  alpha=g@fit$coef[[3]]
  beta=g@fit$coef[[4]]
  mu=mean(goog$Log.Return[2:i])
  sigma=sd(goog$Log.Return[2:i])
  temp.sigma.hat.sq=omega+alpha*mu^2+beta*sigma^2
  goog$sigma.hat.square[i+1]=temp.sigma.hat.sq
}
```

```{r Compute_sigma_hat,echo=TRUE,eval=TRUE}
#compute the value of sigma hat
goog$sigma.hat=sqrt(goog$sigma.hat.square)
```

```{r Compute_r_hat_and_plot,echo=TRUE,fig.align="center",eval=TRUE}
#compute the r hat, and plot the back testing log return value with the r hat value (predicted value)
goog$r.hat=goog$mu.hat+goog$sigma.hat*rnorm(1,0,1)
plot(goog$Log.Return[2801:l],type="l",lty=3,col=1,ylim=c(-0.06,0.08),ylab="")
par(new=TRUE)
plot(goog$r.hat[2801:l],type="l",lty=1,col=2,ylim=c(-0.06,0.08),ylab="")
legend("topright",legend=c("original data","predicted data"),lty=c(3,1),col=c(1,2))
```

</font>

#<font face="Baskerville Old Face">Model: Markov Switching - AR(1) - GARCH(1,1)</font>

<font face="Times New Roman" size=5>

In this model, the `Log.Return` was divided into two groups: one with relative high volatility and one with relative low volatility.

Method 1: Set a pair of threshold level, $n_1$ and $n_2$, where $n_1>n_2$. In `Log.Return` data, any value greater than $n_1$ or less than $n_2$ would be put into group `High`, and the left would be put into group `Low`.

```{r Creat_High_Group_And_Low_Group,echo=TRUE,fig.align="center"}
#find the thresholds, plot out the graph
n1=mean(goog$Log.Return[2:l])+0.02
n2=mean(goog$Log.Return[2:l])-0.02
count=0
while (count<l*0.4){
  count=0
  for (i in 2:l){
    if (goog$Log.Return[i]>n1|goog$Log.Return[i]<n2){
      count=count+1
    }
  }
  if (count<l*0.4){
    n1=n1-0.001
    n2=n2+0.001
  }
}
print(count)
plot(goog$Log.Return[2:l],type="l",lty=3,ylab="Log Return",main="Alphabet Inc. Log Return (Grouping)")
abline(h=n1,col="red",lty=1)
abline(h=n2,col="blue",lty=1)
legend("topright",legend=c("Log Return","n1","n2"),col=c("black","red","blue"),lty=c(3,1,1))
#get the data of group high and group low
High=c()
Low=c()
for (i in 2:l){
  if (goog$Log.Return[i]>n1|goog$Log.Return[i]<n2){
    High=c(High,goog$Log.Return[i])
  }
  else{
    Low=c(Low,goog$Log.Return[i])
  }
}
```

</font>