---
title: "Lab 4 - Simulation from chisquared, t and F-distribution"
output:
  pdf_document: default
  html_document: default
  word_document: default
theme: cerulean
html_document:
  highlight: pygments
word_document: default
---
  -----
  
Here we will just go over the two problems posed in the lab. 


First, using the Grass example:

```{r,echo=TRUE}
Grass = read.table('Grass.csv',head=TRUE,sep=',')
Grass$Region = as.factor(Grass$Region)
```
In these data, we have coded Cultivar with two indicator functions, but left Region as a factor in R. Our basic model is 
```{r,echo=TRUE}
mod = lm(Speed~.,data=Grass)
summary(mod)
```



**Problem** construct a test for Region. 


Here we will make the observation that changes in sums of squares can also be obtained from a change in sum of squared errors. That is
$$
H - H_{-j} = (I-H_{-j}) - (I-H)
$$
So in this case we'll take a shortcut and look at a model without region
```{r,echo=TRUE}
mod2 = lm(Speed~.-Region,data=Grass)
```
and we can now look at the difference between fitted values
```{r,echo=TRUE}
SS.R = sum( mod2$residuals^2 ) - sum( mod$residuals^2 )
SS.R
```
We still need to get a mean-square, given by the 7 region effects we drop. And we can now plug in MSE from the full model, too.
```{r,echo=TRUE}
MS.R = SS.R/7
sighat2 = (summary(mod)$sigma)^2
```
Now we can obtain the $F$ statistic which we test with
```{r,echo=TRUE}
F.R = MS.R/sighat2
1-pf(F.R,7,mod$df.residual)
```
Lets see how this compares to 
```{r,echo=TRUE}
library(car)
Anova(mod)
```

**Problem** perform an experiment to calculate the number of data points necessary to achieve 80% power to find the effect of C2 and C3.  To do this 
   a)  assume that for a given $n$, SS.C is given by $n K$ where you estimate $K$ from the data by $SS.C/n$ using our SS.C above. 
   b) remember that the denominator degrees of freedom change with n, but the numerator remains the same at 2. 
   
Plot the power as n increases (what is the minimum n you could have?) and find where it crosses 80%. 

*Here we need to re-create SS.C (since I'm in a separate file)*
```{r,echo=TRUE}
mod3 = lm(Speed~.-C2-C3,data=Grass)
SS.C = sum(mod3$residuals^2) - sum(mod$residuals^2)
K = SS.C/nrow(Grass)
```
*Let's first look at power with just 13 observations. That is, we have a threshold  $F^{2,1-\alpha}_2$ (since we use 11 degrees of freedom on the model) and look at the probability that a non-central $F^2_2(K)$ exceeds it*
```{r,echo=TRUE}
F.crit = qf(0.95,2,2)
1-pf(F.crit,2,2,K)
```
*We're barely over 0.05, but we'd like to get to 0.8.  So here let's try up to 30, remembering that the denominator degrees of freedom is always n-11.  To do that we'll set up
```{r,echo=TRUE}
ns = 13:30  # potential number of observations
F.crit = qf(0.95,2, ns - 11)  # Sequence of critical values
plot(ns,F.crit)
```
*where the critical value reduces towards about 3.*

*Now our non-centrality parameter also increases: its $nK$ as well as the degrees of freedom for the alternative distribution changing. This gives us*
```{r,echo=TRUE}
p.Fs = 1-pf(F.crit, 2, ns-11,ncp = ns*K)
plot(ns,p.Fs)
abline(h= 0.80)
p.Fs
```
*Where we see that the first time we cross 80% is at 23 observations.*